\section{Introduction}

Vision-language-action (VLA) policies promise general-purpose robot
manipulation, but reliable manipulation requires more than recognizing the
object named by language. Recent robot foundation models scale visual-language
pretraining and robot demonstrations to produce policies that transfer across
tasks and embodiments~\citep{brohan2022rt1,brohan2023rt2,team2024octo,kim2024openvla}.
For closed-loop control, however, the policy must also know where its own body
is in the image: actions are executed by a gripper or end-effector whose
location must be related to proprioceptive state, task objects, and camera
geometry. This makes self-grounding, the binding between proprioception and
visual evidence of the acting body, a central requirement for spatially
reliable VLA control.

We make this requirement concrete with two diagnostic findings. First, when the
physical robot-scene state is fixed and only the camera viewpoint is laterally
shifted, VLA-predicted trajectories shift in the same image direction. Second,
even robot-centric prompts such as ``Where is the robot itself?'' often produce
attention maps that fail to localize the gripper or end-effector. These
findings reveal a gap between task grounding and self-grounding: a policy can
identify the instructed object while failing to bind its action prediction to
the visible body that will execute the action.

This missing-self failure suggests that current VLA adaptation can leave the
action frame entangled with the image frame. A policy trained only with
demonstration action loss under a fixed observation distribution can exploit
pixel-coordinate shortcuts, camera-specific correlations, or object-location
priors that are predictive in the training data. Proprioception may be present
as an input, but it need not constrain the visual representation to encode
where the robot appears. As a result, an otherwise competent policy can remain
sensitive to camera-only perturbations because its action-relevant visual
features are not anchored to the robot's own embodiment.

Existing spatial and grounding-oriented VLA methods address important parts of
this problem, but they do not directly supply a self-supervised self-grounding
signal. Recent methods improve spatial reasoning through visual trajectory
prompts, 3D representations, camera-space action formulations, view synthesis,
or explicit treatments of proprioception~\citep{zheng2025tracevla,qu2025spatialvla,zhang2025ocvla,gu2026vistabot,zhao2025statefree,lu2026gap}.
These directions make space, viewpoint, or state more explicit, yet they
typically supervise task execution rather than the visual binding between the
acting body and proprioceptive change. What is missing is a training signal
that tells the VLA where its own visible embodiment is, without requiring
manual masks, external robot-part annotations, or stronger teacher models.

\textsc{ProprioVLA} addresses this gap by turning robot interaction into
self-grounding supervision. The key insight is that the robot reveals itself
through visual change that is synchronized with proprioceptive change: when the
end-effector moves or the gripper changes, the corresponding image evidence is
not arbitrary background motion but evidence of the body that acts. We convert
temporal image differences, optical flow, and proprioceptive activity into
token-level self-grounding targets over the VLA visual tokens. These targets
provide an implicit answer to the question ``where is the robot itself?''
during training, while requiring only the observations and states already
available in robot demonstrations.

The method has two complementary components. State-Aligned Residual Flow (SARF)
mines confidence-weighted self-grounding targets by selecting visual change
that is locally salient, temporally coherent, and supported by proprioceptive
activity. A Proprioceptive Alignment Module (PAM) then trains a lightweight
learned-query grounding head over the VLA visual representation to recover
these targets with normalized KL supervision. Importantly, PAM is an auxiliary
side head: it does not concatenate pseudo-labels or grounding latents into the
action path, and the mined targets are not needed at inference. Instead, the
auxiliary loss regularizes shared visual tokens to encode the visible
embodiment while preserving the standard action-supervised VLA objective.

Experiments test whether this self-grounding signal improves spatial
reliability beyond ordinary fine-tuning and state concatenation. We evaluate
\textsc{ProprioVLA} in simulation and real-robot settings, including controlled
camera-perturbation probes that isolate view-induced action drift. Across these
settings, \textsc{ProprioVLA} improves task success and robustness over
action-supervised VLA adaptation. Ablations further isolate the design choices:
state-aligned target mining is better than raw motion saliency alone, and
normalized KL supervision provides a stable objective for learned-query
self-grounding across token counts and target sharpness.

Our contributions are:
\begin{itemize}
    \item We identify a missing-self failure mode in VLA policies: camera-only
    viewpoint shifts can induce co-directional shifts in predicted robot
    trajectories, while robot-centric prompting can still fail to localize the
    gripper or end-effector.
    \item We introduce State-Aligned Residual Flow (SARF), which turns
    proprioceptively aligned optical-flow and image-difference cues into
    confidence-weighted token-level self-grounding targets without manual
    masks, external grounding datasets, or stronger foundation-model
    supervision.
    \item We propose \textsc{ProprioVLA}, a self-supervised proprioceptive
    alignment approach that trains a learned-query grounding head with
    normalized KL supervision, improving task success and robustness in
    real-robot and simulation environments.
\end{itemize}
