\section{\textsc{ProprioVLA}}

\textsc{ProprioVLA} adds a proprioceptive grounding path to a VLA policy while
leaving the standard action-supervised objective intact. During training, the
method mines self-grounding targets from state-aligned residual motion, uses
proprioceptive state as a query over visual features, and anchors the queried
visual representation by reconstructing proprioceptive information. At test
time, only the learned state-conditioned attention path is used; the mined
heatmaps are not required.

\subsection{Self-Grounding from State-Aligned Motion}

The robot reveals itself through motion that is synchronized with its state
change. Given consecutive observations \((o_t,o_{t+1})\) and proprioceptive
states \((s_t,s_{t+1})\), we define
\(\Delta s_t=s_{t+1}-s_t\). We estimate dense visual motion \(F_t(p)\) between
frames and subtract global or background motion to obtain residual motion
\(R_t(p)\) at image location \(p\). This residual suppresses camera drift and
static-scene variation, emphasizing motion caused by the robot or by objects
directly coupled to the robot.

We convert residual motion into a confidence-weighted self-grounding target.
Let \(\mathrm{sal}(R_t(p))\) denote local residual-motion saliency and let
\[
    \alpha_t(p)
    =
    \left[
    \cos\bigl(\psi_R(R_t(p)),\psi_s(\Delta s_t)\bigr)
    \right]_+
\]
measure state-motion alignment after lightweight projections \(\psi_R\) and
\(\psi_s\). The unnormalized heatmap score is
\[
    e_t(p)=\mathrm{sal}(R_t(p))\,\alpha_t(p).
\]
We normalize \(e_t\) over the image or patch grid to obtain \(H_t\), and set
a confidence \(c_t\in[0,1]\) from the total aligned-motion mass. Low confidence
downweights frames where the self-motion cue is weak or ambiguous. Thus the
training target is mined from interaction alone, without manual masks, external
grounding datasets, or stronger foundation-model supervision.

\subsection{Proprioception as a Visual Query}

In \textsc{ProprioVLA}, proprioception determines where the policy looks rather
than merely adding another token to the transformer. We encode the
proprioceptive state as
\[
    q_t^s=f_s(s_t),
\]
and obtain visual keys and values \(K_t,V_t=f_v(o_t)\) from the VLA vision
encoder. The state query attends over visual patches according to
\[
    A_t
    =
    \mathrm{softmax}
    \left(
    \frac{q_t^s K_t^\top}{\sqrt{d}}
    \right),
    \qquad
    z_t^{\mathrm{self}} = A_t V_t .
\]
Here \(A_t\) is a distribution over image patches, and
\(z_t^{\mathrm{self}}\) is the self-grounded visual feature passed to the
action head together with the policy's standard visual-language representation.

The mined heatmap supervises this state-conditioned attention through
\[
    \mathcal{L}_{\mathrm{ground}}
    =
    c_t\,D(A_t,H_t),
\]
where \(D\) is cross-entropy or KL divergence after resizing \(H_t\) to the
attention grid. This loss turns proprioception into an active grounding
mechanism: the state embedding selects image evidence corresponding to the
current state. At inference time, the learned query directly produces \(A_t\)
and \(z_t^{\mathrm{self}}\) from the current observation and state.

\subsection{Proprioceptive Anchoring by Reconstruction}

The self-grounding loss specifies where to attend, but it does not guarantee
what the attended feature encodes. We therefore add a reconstructive
proprioceptive anchoring objective. The reconstruction path is
\[
    s_t \rightarrow q_t^s \rightarrow A_t \rightarrow
    z_t^{\mathrm{self}} \rightarrow \hat{s}_t ,
\]
where the attended representation is explicitly
\[
    z_t^{\mathrm{self}}=\sum_p A_t(p)V_t(p),
    \qquad
    \hat{s}_t=g_s(z_t^{\mathrm{self}}).
\]
The state reconstruction loss is
\[
    \mathcal{L}_{\mathrm{state}}
    =
    \|\hat{s}_t-s_t\|_2^2 .
\]
For paired observations, we also reconstruct state change:
\[
    \widehat{\Delta s_t}
    =
    g_\Delta(z_t^{\mathrm{self}},z_{t+1}^{\mathrm{self}}),
    \qquad
    \mathcal{L}_{\mathrm{motion}}
    =
    \|\widehat{\Delta s_t}-\Delta s_t\|_2^2 .
\]
The full reconstruction objective is
\[
    \mathcal{L}_{\mathrm{rec}}
    =
    \mathcal{L}_{\mathrm{state}}
    +
    \beta \mathcal{L}_{\mathrm{motion}},
\]
and the complete training loss is
\[
    \mathcal{L}
    =
    \mathcal{L}_{\mathrm{act}}
    +
    \lambda_g \mathcal{L}_{\mathrm{ground}}
    +
    \lambda_r \mathcal{L}_{\mathrm{rec}} .
\]
Here \(\mathcal{L}_{\mathrm{act}}\) is the standard imitation loss used to
fine-tune the VLA. The reconstruction heads receive only
\(z_t^{\mathrm{self}}\), never the raw state \(s_t\) or state embedding
\(q_t^s\). This implementation detail prevents trivial copying and makes
reconstruction a visual anchoring regularizer. The policy is therefore trained
not only to look at the robot, but also to encode the robot's state in the
visual feature it attends to.
