\section{Related Work}

\paragraph{Vision-language-action models.}
Vision-language-action (VLA) policies scale language-conditioned robot control
by coupling vision-language backbones with robot action prediction. RT-1 and
RT-2 established the value of large robot datasets and web-scale semantic
priors~\citep{brohan2022rt1,brohan2023rt2}; recent open and generalist models
further improve transfer across tasks, embodiments, and action
parameterizations~\citep{team2024octo,kim2024openvla,black2024pi0}.
However, this line of work mainly optimizes task success under the training
observation distribution. It does not directly test whether the action-relevant
visual evidence is anchored to camera-invariant robot-object geometry.
\textsc{ProprioVLA} studies this missing failure mode: under camera
perturbations, a fine-tuned VLA can follow image-coordinate shortcuts instead
of grounding actions in the robot's own state.

\paragraph{Spatial and proprioceptive grounding in VLA policies.}
Recent work improves VLA perception by making space, viewpoint, or
proprioception more explicit. TraceVLA adds visual trajectory prompts to
increase spatial-temporal awareness~\citep{zheng2025tracevla}; SpatialVLA
injects 3D spatial representations~\citep{qu2025spatialvla}; OC-VLA predicts
actions in camera space to reduce observation-action mismatch~\citep{zhang2025ocvla};
and VistaBot uses geometry-aware view synthesis for cross-view
robustness~\citep{gu2026vistabot}. These methods show that perception is a
central bottleneck, but they depend on explicit prompts, 3D encodings, camera
coordinate transforms, or generative view modules. A complementary line
questions proprioception itself: State-free Policy removes proprioceptive
inputs to avoid trajectory overfitting~\citep{zhao2025statefree}, while GAP
shows that proprioceptive gradients can suppress visual learning during motion
transitions~\citep{lu2026gap}. \textsc{ProprioVLA} takes a different stance: it
does not discard proprioception or treat it as a passive token. Instead, it
uses proprioceptive change to mine state-aligned residual-flow heatmaps, turning
the robot's own motion into self-supervision for learned visual attention.
